### 2020-03-25 thesis/content/chapter-related-work.tex
complted gcnn lta relation section

\subsubsection{Spatial \acp{gcnn}}
Looking at the \ac{gcn} layer defined in \cref{eq:related:gcn-layer}, notice that, even though it is motivated by spectral graph theory, the spectrum $\lambda_G$ is not actually directly used.
This allows for a very different perspective on \acp{gcn}:
Instead of interpreting convolutions as applications of linear spectral filters, we can interpret them as vertex neighborhood aggregations, similar to 1-\acs{wl}.
From this perspective a \ac{gcn} can be seen as a so-called spatial \ac{gcnn} because it operates directly in the vertex domain.
To see why this is the case, we rewrite \cref{eq:related:gcn-layer} and compare it with the 1-\acs{wl} color refinement strategy (see \cfullref{defn:related:wl1-refine}):
\begin{align*}
	\text{\ac*{gcn}:}\quad&& Z^{(t)}[v] &= \sigma\left(\left( {\mu(v)}^2 Z^{(t-1)}[v] + \smashoperator[lr]{\sum_{u \in \Gamma_G(v)}} \mu(v) \mu(u) Z^{(t-1)}[u] \right) \Theta^{(t)} \right) \\
	\text{1-\acs{wl}:}\quad&& \chi_{G,1}^{(t)}(v) &= h\left( \chi_{G,1}^{(t-1)}(v), \ldblbrace \chi_{G,1}^{(t-1)}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right)
\end{align*}
Here $\mu(v) \coloneqq {\left(\left| \Gamma_G(v) \right| + 1\right)}^{-\frac{1}{2}}$ are normalization factors introduced by $\tilde{D}_G^{-\frac{1}{2}}$.
The comparison shows that \ac{gcn} convolutions can be understood as iterative refinements of continuous ``color vectors'' $Z^{(t)}[v] \in \mathbb{R}^{d^{(t)}}$.
Then the continuous analogue to the injective 1-\acs{wl} hash function $h: \mathcal{C}^{*} \to \mathcal{C}$ can be defined as $h_{\text{\acs*{gcn}}}(z_0, z_1, \dots, z_m) = \sigma(\sum_{j=0}^m \mu(v_0) \mu(v_j) z_j \Theta^{(t)})$ which computes a single color vector from multiple color vectors $z_j$ where $v_j$ are the corresponding vertices.

@@

The graph convolution approaches that we just looked at all produce a set of convolved feature vectors ${\left\{ Z^{(T)}[s_i] \right\}}_{i=1}^{m}$ after $T$ convolutional layers. % chktex 21
For 1-\acs{wl} bounded architectures like \ac{gcn} or \ac{gin}, there is one vector for each of the $m = {|\mathcal{V}_G|}$ vertices, while a $k$-\acs{wl} inspired architecture like $k$-\acs{gnn} produces $m \leq {|\mathcal{V}_G|}^{k}$ output vectors.
In order to solve the \ac{gcr} problem, those vector sets need to be combined into a final predicted class or regression value.
To do this, so-called \textit{graph pooling layers} are used.
Generally speaking there are two types of pooling approaches:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Hierarchical Pooling:}
		This type of pooling is similar to that found in \acp{cnn} on pictures where the resolution is iteratively reduced.
		Similarly hierarchical graph pooling iteratively \textit{coarsens} a graph every couple of convolutional layers.
		Graph coarsening works by merging vertices in a spectrum-preserving manner~\cite{Loukas2018}.
		This is done until a single merged vertex is left whose feature vector then represents the entire graph.
	\item \textbf{Global Pooling}:
		Alternatively pooling can also be performed in a single step after the convolutional layers.
		This type of pooling takes all $m$ feature vectors and directly maps them to a single graph feature vector.
\end{enumerate}
In this thesis we will focus on global pooling.
The most simple global pooling approaches use static aggregation functions like component-wise $\min$, $\max$ or $\mean$ to produce a single graph feature vector $z_G \in \mathbb{R}^{d^{(L)}}$.
This vector is then fed into a standard \ac{mlp} to produce the final prediction.
The state-of-the-art graph pooling approaches go beyond static aggregation and try to incorporate structural information.
We will now briefly look at two such approaches.

\paragraph{SortPooling}
One way to combine feature vectors is called \textit{SortPooling}~\cite{Zhang2018}.
It interprets the convolved vectors as continuous \ac{wl} colors that encode different structural roles of vertices (or vertex sets in case of $k$-\acp{gnn}).
By imposing a component-wise lexicographic ordering on the set of color vectors, it can be reduced to a vertex permutation invariant top-$p$ list of vectors $(z_1, \dots, z_p)$ for some fixed $p \in \mathbb{N}$.
The final graph feature vector then is the concatenation $z_G = \left( \bigoplus_{i=1}^{p} z_i \right) \in \mathbb{R}^{d^{(L)} p}$.
SortPooling is based on the idea that the convolutional layers will learn to use the most-significant lexicographic vector components to represent that vector's importance.

>>>>

\subsubsection{Spatial \acp{gcnn}}
Looking at the \ac{gcn} layer defined in \cref{eq:related:gcn-layer}, notice that, even though it is motivated by spectral graph theory, the spectrum $\lambda_G$ is not actually directly used.
This allows for a very different perspective on \acp{gcn}:
Instead of interpreting convolutions as applications of linear spectral filters, we can interpret them as vertex neighborhood aggregations, similar to 1-\acs{wl}.
From this perspective a \ac{gcn} can be seen as a so-called spatial \ac{gcnn} because it operates directly in the vertex domain.
To see why this is the case, we rewrite \cref{eq:related:gcn-layer} and compare it with the 1-\acs{wl} color refinement strategy (see \cfullref{defn:related:wl1-refine}):
\begin{alignat*}{2}
	\text{\ac*{gcn}:}\quad&& Z^{(t)}[v] &= \sigma\left(\left( {\mu(v)}^2 Z^{(t-1)}[v] + \smashoperator[lr]{\sum_{u \in \Gamma_G(v)}} \mu(v) \mu(u) Z^{(t-1)}[u] \right) \Theta^{(t)} \right) \\
	\text{1-\acs{wl}:}\quad&& \chi_{G,1}^{(t)}(v) &= h\left( \chi_{G,1}^{(t-1)}(v), \ldblbrace \chi_{G,1}^{(t-1)}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right)
\end{alignat*}
Here $\mu(v) \coloneqq {\left(\left| \Gamma_G(v) \right| + 1\right)}^{-\frac{1}{2}}$ are normalization factors introduced by $\tilde{D}_G^{-\frac{1}{2}}$.
The comparison shows that \ac{gcn} convolutions can be understood as iterative refinements of continuous ``color vectors'' $Z^{(t)}[v] \in \mathbb{R}^{d^{(t)}}$.
Then the continuous analogue to the injective 1-\acs{wl} hash function $h: \mathcal{C}^{*} \to \mathcal{C}$ can be defined as $h_{\text{\acs*{gcn}}}(z_0, z_1, \dots, z_m) = \sigma(\sum_{j=0}^m \mu(v_0) \mu(v_j) z_j \Theta^{(t)})$ which computes a single color vector from multiple color vectors $z_j$ where $v_j$ are the corresponding vertices.

@@

The graph convolution approaches that we just looked at all produce a set of convolved feature vectors ${\left\{ Z^{(T)}[s_i] \right\}}_{i=1}^{m}$ after $T$ convolutional layers. % chktex 21
For 1-\acs{wl} bounded architectures like \ac{gcn} or \ac{gin}, there is one vector for each of the $m = {|\mathcal{V}_G|}$ vertices, while a $k$-\acs{wl} inspired architecture like $k$-\acs{gnn} produces $m \leq {|\mathcal{V}_G|}^{k}$ output vectors.
In order to solve the \ac{gcr} problem, those vector sets need to be combined into a final predicted class or regression value.
To do this, so-called \textit{graph pooling layers} are used.
Generally speaking there are two types of pooling approaches:
\begin{enumerate}[label=\textbf{\arabic*.}]
	\item \textbf{Hierarchical Pooling:}
		This type of pooling is similar to that found in \acp{cnn} on pictures where the resolution is iteratively reduced.
		Similarly hierarchical graph pooling iteratively \textit{coarsens} a graph every couple of convolutional layers.
		Graph coarsening works by merging vertices in a spectrum-preserving manner~\cite{Loukas2018}.
		This is done until a single merged vertex is left whose feature vector then represents the entire graph.
	\item \textbf{Global Pooling}:
		Alternatively pooling can also be performed in a single step after the convolutional layers.
		This type of pooling takes all $m$ feature vectors and directly maps them to a single graph feature vector.
\end{enumerate}
In this thesis we will focus on global pooling.
The most simple global pooling approaches use static aggregation functions like component-wise $\min$, $\max$ or $\mean$ to produce a single graph feature vector $z_G \in \mathbb{R}^{d^{(T)}}$.
This vector is then fed into a standard \ac{mlp} to produce the final prediction.
The state-of-the-art graph pooling approaches go beyond static aggregation and try to incorporate structural information.
We will now briefly look at two such approaches.

\paragraph{SortPooling}
One way to combine feature vectors is called \textit{SortPooling}~\cite{Zhang2018}.
It interprets the convolved vectors as continuous \ac{wl} colors that encode different structural roles of vertices (or vertex sets in case of $k$-\acp{gnn}).
By imposing a component-wise lexicographic ordering on the set of color vectors, it can be reduced to a vertex permutation invariant top-$p$ list of vectors $(z_1, \dots, z_p)$ for some fixed $p \in \mathbb{N}$.
The final graph feature vector then is the concatenation $z_G = \left( \bigoplus_{i=1}^{p} z_i \right) \in \mathbb{R}^{d^{(T)} p}$.
SortPooling is based on the idea that the convolutional layers will learn to use the most-significant lexicographic vector components to represent that vector's importance.
